Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2022-07-29 paperDraft/paperDraft4.Rmd
Updates draft

- akapit filozoficzny, z odniesieniem do value free

Modern Natural Language Processing (NLP) models are used to complete various tasks such as providing email filters, smart assistants, search results, language translations, text analytics and so on. All of them need as an input words represented by means of numbers which is accomplished with word embeddings, in which particular lexical units are represented as vectors of real numbers.

- troche o tym jak sa skonstruowane (optimizing for co-ocurrence frequency)

- przejrzec jak to jest wprowadzane w kilku innych artykulach, zwl. fair is better than sensational

- troche wiecej o tej konstrukcji i intuicji semantycznych, ale to nie super dobry pomysl

- undesirable bias with respect to a certain groups words if stereotypically connected words (for stereotypes that we don't want to be used or relied on in downstream tasks) are systematically closer to each other. + EXAMPLES

It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that in the learning process  such models can learn implicit biases that reflect harmful stereotypical thinking. A large chunk of  the literature on the topic  focuses on  the geometry of the learnt word embeddings;  in particular, unusually low (cosine) proximity of words belonging, intuitively, to a stereotype, is taken as a sign that  this stereotype has been built into a given embedding. This methodology will be in the focus of our paper.

>>>>

Modern Natural Language Processing (NLP) models are used to complete various tasks such as providing email filters, smart assistants, search results, language translations, text analytics and so on. All of them need as an input words represented by means of numbers which is accomplished with word embeddings, in which particular lexical units are represented as vectors of real numbers. They are necessary to perform machine learning tasks that require the use of natural language. One of the ways to learn these vectors representations is to use simple neural network that based on the millions of real sentences examples assigns the numbers based on their co-occurrence and context. Using co-occurence results in a vector space that contains similar words closer to each other and dissimilar further away. It has been suggested [@Bolukbasi2016man; @Caliskan2017semanticsBiases; @Gonen2019lipstick; @Lauscher2019multidimensional; @Garg2018years; @Manzini2019blackToCriminal]  that in the learning process  such models can learn implicit biases that reflect harmful stereotypical thinking (eg. that the word \textit{she} is much closer in the vector space to the word \textit{cooking} than the word \textit{he}). This phenomenon may be undesired when using these vectors for further downstream tasks (eg. web search, recommendations, etc.).\todo{find downstream tasks and bias examples from papers} Before we move to more technical details regarding this topic, it is worth looking first at the theoretical background which may help to better grasp the concept of bias within machine learning algorithms.

In \todo{cite Are algorithms Value-Free} the author focuses on the general issue, namely whether algorithms can in principle be value-free. The author draws attention to the observation that in both scientific research and machine learning domain, induction plays a crucial role in the decision-making process. 
Induction is essential for both of these domains as "if ever considered literally just the facts, we would never be able to draw inductive conclusions" \todo{cite Gabrielle}. Induction comes with certain risks which are 1) the problem with justification, and 2) the risk of getting things wrong. During the process of designing machine learning process some decisions have to always be made. As the author states, there is no such thing as the justification in the abstract. \todo{cite Gabrielle} We may justify certain design decisions, data choice, etc. only by referring to the world's being certain way. Therefore if the induction is justified by contingent features that the world exhibits, which are intrinsically dependent on our current set of values, then our values indeed are the source of the justified inductive arguments. The other issue with induction is that as the machine learning algorithm always has access to a limited amount of information, it is prone to make mistakes from time to time, even if it's general accuracy is very high. What does it all mean in the context of bias within word embeddings? It seems that the presence of bias, value-laden algorithms is unavoidable due to the very nature of induction.

Assuming that the algorithms indeed due to the nature of their decision-making cannot fulfill the ideal of being value-free, it is even more justified to focus on ways how to properly detect harmful bias to limit its negative impact by the awareness of its presence. The aim of this paper is to contribute to this idea that using proper statistical tools may improve our understanding of what is happening inside machine learning models and especially within word embeddings.

### 2022-07-29 paperDraft/paperDraft4.tex
Updates draft

  dvipsnames,enabledeprecatedfontcommands, twocolumn]{scrartcl}

@@

\begin{itemize}
\tightlist
\item
  akapit filozoficzny, z odniesieniem do value free
\end{itemize}

Modern Natural Language Processing (NLP) models are used to complete
various tasks such as providing email filters, smart assistants, search
results, language translations, text analytics and so on. All of them
need as an input words represented by means of numbers which is
accomplished with word embeddings, in which particular lexical units are
represented as vectors of real numbers.

\begin{itemize}
\item
  troche o tym jak sa skonstruowane (optimizing for co-ocurrence
  frequency)
\item
  przejrzec jak to jest wprowadzane w kilku innych artykulach, zwl. fair
  is better than sensational
\item
  troche wiecej o tej konstrukcji i intuicji semantycznych, ale to nie
  super dobry pomysl
\item
  undesirable bias with respect to a certain groups words if
  stereotypically connected words (for stereotypes that we don't want to
  be used or relied on in downstream tasks) are systematically closer to
  each other. + EXAMPLES
\end{itemize}

It has been suggested {[}1--6{]} that in the learning process such
models can learn implicit biases that reflect harmful stereotypical
thinking. A large chunk of the literature on the topic focuses on the
geometry of the learnt word embeddings; in particular, unusually low
(cosine) proximity of words belonging, intuitively, to a stereotype, is
taken as a sign that this stereotype has been built into a given
embedding. This methodology will be in the focus of our paper.

@@

\hypertarget{weat}{%
\subsection{WEAT}\label{weat}}

\hypertarget{mac}{%
\subsection{MAC}\label{mac}}

>>>>

  dvipsnames,enabledeprecatedfontcommands]{scrartcl}

@@

Modern Natural Language Processing (NLP) models are used to complete
various tasks such as providing email filters, smart assistants, search
results, language translations, text analytics and so on. All of them
need as an input words represented by means of numbers which is
accomplished with word embeddings, in which particular lexical units are
represented as vectors of real numbers. They are necessary to perform
machine learning tasks that require the use of natural language. One of
the ways to learn these vectors representations is to use simple neural
network that based on the millions of real sentences examples assigns
the numbers based on their co-occurrence and context. Using co-occurence
results in a vector space that contains similar words closer to each
other and dissimilar further away. It has been suggested {[}1--6{]} that
in the learning process such models can learn implicit biases that
reflect harmful stereotypical thinking (eg. that the word \textit{she}
is much closer in the vector space to the word \textit{cooking} than the
word \textit{he}). This phenomenon may be undesired when using these
vectors for further downstream tasks (eg. web search, recommendations,
etc.).\todo{find downstream tasks and bias examples from papers} Before
we move to more technical details regarding this topic, it is worth
looking first at the theoretical background which may help to better
grasp the concept of bias within machine learning algorithms.

In \todo{cite Are algorithms Value-Free} the author focuses on the
general issue, namely whether algorithms can in principle be value-free.
The author draws attention to the observation that in both scientific
research and machine learning domain, induction plays a crucial role in
the decision-making process. Induction is essential for both of these
domains as ``if ever considered literally just the facts, we would never
be able to draw inductive conclusions'' \todo{cite Gabrielle}. Induction
comes with certain risks which are 1) the problem with justification,
and 2) the risk of getting things wrong. During the process of designing
machine learning process some decisions have to always be made. As the
author states, there is no such thing as the justification in the
abstract. \todo{cite Gabrielle} We may justify certain design decisions,
data choice, etc. only by referring to the world's being certain way.
Therefore if the induction is justified by contingent features that the
world exhibits, which are intrinsically dependent on our current set of
values, then our values indeed are the source of the justified inductive
arguments. The other issue with induction is that as the machine
learning algorithm always has access to a limited amount of information,
it is prone to make mistakes from time to time, even if it's general
accuracy is very high. What does it all mean in the context of bias
within word embeddings? It seems that the presence of bias, value-laden
algorithms is unavoidable due to the very nature of induction.

Assuming that the algorithms indeed due to the nature of their
decision-making cannot fulfill the ideal of being value-free, it is even
more justified to focus on ways how to properly detect harmful bias to
limit its negative impact by the awareness of its presence. The aim of
this paper is to contribute to this idea that using proper statistical
tools may improve our understanding of what is happening inside machine
learning models and especially within word embeddings.

@@

\vspace{1mm}
\footnotesize

\begin{figure}

\begin{center}\includegraphics[width=1.1\linewidth]{paperDraft4_files/figure-latex/unnamed-chunk-2-1} \end{center}
\caption{dsds}
\label{fig:weat7google}
\end{figure}
